---
title       : Intro to R Programming for Biostatistics
subtitle    : "Day 2 - Arranging and Summarizing Data in R"
author      : Adam J Sullivan
job         : 
license     : by-nc-nd
framework   : io2012        # {io2012, html5slides, shower, dzslides, ...}
highlighter : highlight.js  # {highlight.js, prettify, highlight}
hitheme     : tomorrow      # 
widgets     : [mathjax, quiz, bootstrap, interactive]            # {mathjax, quiz, bootstrap}
mode        : selfcontained # {standalone, draft}
logo        : shield_image.png
biglogo     : shield_image_large.png
knit        : slidify::knit2slides
assets      : {assets: ../../assets}
---  .segue bg:grey

```{r setup, include = FALSE, cache = FALSE}
knitr::opts_chunk$set(error = TRUE)
knitr::opts_chunk$set(cache=TRUE)
knitr::opts_chunk$set(warning=FALSE)
knitr::opts_chunk$set(message=FALSE)
```

```{r, echo=F}
require(nycflights13)
require(tidyverse)
```

# Arranging the Data

--- .class #id

## Arranging the Data

- We also have need to make sure the data is ordered in a certain manner. This can be easily done in R with the `arrange()` function. 
- Again we can do this in base R but this is not always a clear path. 

--- .class #id

## Arranging the Data Example

- Let's say that we wish to look at only carriers and departure delay and we wish to order departure delays from what smallest to largest. 
- In base R we would have to run the following command:

```{r results='hide'}
flights[order(flights$dep_delay), c("carrier", "dep_delay")]
```

--- .class #id

## Enter the `arrange()` Function

We could do this in an easy manner using the  `arrange()` function:

```
arrange(.data, ...)
```

Where

- `.data` is a data frame of interest.
- `...` are the variables you wish to sort by. 

--- .class #id

## Arranging the Data Example Continued

```{r, eval=FALSE}

flights %>%
    select(carrier, dep_delay) %>%
    arrange(dep_delay)
```

--- .class #id

## Arranging the Data Example Continued

```{r, echo=FALSE}

flights %>%
    select(carrier, dep_delay) %>%
    arrange(dep_delay)
```

--- .class #id

## Arranging the Data Example Continued

- With `arrange()` we first use `select()` to pick the only columns that we want and then we arrange by the `dep_delay`. 
- If we had wished to order them in a descending manner we could have simply used the `desc()` function:

```{r, eval=FALSE }
flights %>%
    select(carrier, dep_delay) %>%
    arrange(desc(dep_delay))
```

--- .class #id


## More Complex Arrange

- Lets consider that we wish to look at the top 3 departure delays for each day.
- Then we wish to order them from largest to smallest departure delay. 
- We then need to do the following:
    1. Group by month and Day
    2. Pick the top 3 departure delays
    3. order them largest to smallest
    



--- .class #id

## More Complex Arrange Continued


```{R, eval=FALSE}
flights %>% 
  group_by(month, day) %>%  
  top_n(3, dep_delay) %>% 
  arrange(desc(dep_delay))
```

Where

- `group_by()` is a way to group data. This way we perform operations on a group. So top 3 delays are by a group of day and month. 
- `top_n()`takes a tibble and returns a specific number of rows based on a chosen value. 


--- .class #id

## More Complex Arrange Continued


```{R, echo=FALSE}
flights %>% 
  group_by(month, day) %>%  
  top_n(3, dep_delay) %>% 
  arrange(desc(dep_delay))
```



--- .class #id

## On Your Own: RStudio Practice

- Perform the following operations:
  - Group by month and day. 
  - use `sample_n()` to pick 1 observation per day. 
  - Arrange by longest to smallest departure delay. 



--- .class #id





## On Your Own: RStudio Practice


Your answer ***may*** look like:

```{r, eval=FALSE}
flights %>%
  group_by(month,day) %>%
  sample_m(1) %>%
  arrange(desc(dep_delay))
```

---  .segue bg:grey


# Summarizing Data


--- .class #id

## Summarizing Data

- As you have seen in your own work, being able to summarize information is crucial. 
- We need to be able to take out data and summarize it as well. 
- We will consider doing this using the `summarise()` function. 

--- .class #id

## Summarizing Data

- Like in the rest of these lessons, let's consider what happens when we try to to do this in base R. We will:
  1. Create a table grouped by `dest`.
  2. Summarize each group by taking mean of `arr_delay`.

```{r results='hide'}
head(with(flights, tapply(arr_delay, dest, mean, na.rm=TRUE)))
head(aggregate(arr_delay ~ dest, flights, mean))
```

--- .class #id


## Enter `summarise()` Function

- The `summarise()` function is:

```
summarise(.data, ...)
```

- where
  - `.data` is the tibble of interest.
  - `...` is a list of name paired summary functions
    - Such as:
      - `mean()`
      - `median`
      - `var()`
      - `sd()`
      - `min()`
      - `max()
      - ...

--- .class #id

## Summarizing Data Example


```
flights %>%
    group_by(dest) %>%
    summarise(avg_delay = mean(arr_delay, na.rm=TRUE))
```

--- .class #id

## Summarizing Data Example 

- Consider the logic here:
  1. Group flights by destination
  2. Find the average delay of the groups and call it `avg_delay`.
- This is much easier to understand than the  Base R code. 

```{r, echo=FALSE}
flights %>%
    group_by(dest) %>%
    summarise(avg_delay = mean(arr_delay, na.rm=TRUE))
```

--- .class #id


## Another Example

- Lets say that we would like to have more than just the averages but we wish to have the minimum and the maximum departure delays by carrier:

```{r, eval=FALSE}

flights %>%
    group_by(carrier) %>%
    summarise_each(funs(min(., na.rm=TRUE), max(., na.rm=TRUE)), matches("delay"))
```


--- .class #id


## Another Example



```{r, echo=FALSE}

flights %>%
    group_by(carrier) %>%
    summarise_each(funs(min(., na.rm=TRUE), max(., na.rm=TRUE)), matches("delay"))
```

--- .class #id

## On Your Own: RStudio Practice 

- The following is a new function:
  - Helper function `n()` counts the number of rows in a group
- Then for each day:
  - count total flights
  - Sort in descending order. 

--- .class #id

## On Your Own: RStudio Practice 

Your answer should look like:

```{r, echo=FALSE}

flights %>%
    group_by(month, day) %>%
    summarise(flight_count = n()) %>%
    arrange(desc(flight_count))
```


We could also have used what is called the  `tally()` function:

```{r}
flights %>%
    group_by(month, day) %>%
    tally(sort = TRUE)
```

---  .segue bg:grey


# Adding New Variables


--- .class #id

## Adding New Variables

- There is usually no way around needing a new variable in your data. 
- For example, most medical studies have height and weight in them, however many times what a researcher is interested in using is Body Mass Index (BMI). 
- We would need to add BMI in. 


--- .class #id

## Adding New Variables

- Using the `tidyverse` we can add new variables in multiple ways
  - `mutate()`
  - `transmute()`

--- .class #id

## Adding New Variables


With `mutate()` we have

```
mutate(.data, ...)
```

where

- `.data` is your tibble of interest.
- `...` is the name paired with an expression


--- .class #id

## Adding New Variables

Then with `transmute()` we have:

```
transmute(.data, ...)
```

where

- `.data` is your tibble of interest.
- `...` is the name paired with an expression


--- .class #id


## Differences Between `mutate()` and `transmute()`

- There is only one major difference between `mutate()` and `transmutate` and that is what it keeps in your data. 
  - `mutate()`
    - creates a new variable
    - It keeps all existing variables
  - `transmute()`
   - creates a new variable.
   - It only keeps the new variables


--- .class #id

## Example

- Let's say we wish to have a variable called speed. We want to basically do:

\[\text{speed} = \dfrac{\text{distance}}{\text{time}}*60\]


We can first do this with `mutate()`:

```{r, eval=FALSE}
flights %>% 
  select(flight, distance, air_time) %>%
  mutate(speed = distance/air_time*60)
```

--- .class #id

## Example

```{r, echo=FALSE}
flights %>% 
  select(flight, distance, air_time) %>%
  mutate(speed = distance/air_time*60)
```

--- .class #id

## Example


```{r, eval=FALSE}
flights %>%
  select(flight, distance, air_time) %>%
  transmute(speed = distance/air_time*60)
```


--- .class #id

## Example


```{r, eval=FALSE}
flights %>%
  select(flight, distance, air_time) %>%
  transmute(speed = distance/air_time*60)
```

---  .segue bg:grey



#Further Summaries

--- .class #id

## Further Summaries

- We have so far discussed how one could find the basic number summaries:
  - mean
  - median
  - standard deviation
  - variance
  - minimum 
  - maximum
- However there are many more operations that you may wish to do for summarizing data. 
- In fact many of the following examples are excellent choices for working with categorical data which does not always make sense to do the above summaries for. 


--- .class #id

## Further Summaries

- We will consider:
  1. Grouping and Counting
  2. Grouping, Counting and Sorting
  3. Other Groupings
  4. Counting Groups


--- .class #id



## Grouping and Counting


- We have seen the functions `tally()` and `count()`. 
- Both of these can be used for grouping and counting. 
- They also are very concise in how they are called. 


--- .class #id



## Grouping and Counting


- For example if we wished to know how many flights there were by month, we would use `tally()` in this manner: 

```{r, eval=FALSE }
flights %>%
  group_by(month) %>% 
  tally()
```

--- .class #id



## Grouping and Counting

- Where as we could do the same thing with `count()`

```{r, eval=FALSE}
flights %>% 
  count(month)
```

*Notice: `count()` allowed for month to be called inside of it, removing the need for the `group_by()` function. 


--- .class #id



## Grouping, counting and sorting. 

- Both `tally()` and `count()` have an argument called `sort()`. 
- This allows you to go one step further and group by, count and sort at the same time. 
- For `tally()` this would be:

```{r, eval=FALSE}
flights %>% group_by(month) %>% tally(sort=TRUE)
```

--- .class #id

```{r, echo=FALSE}
flights %>% group_by(month) %>% tally(sort=TRUE)
```


--- .class #id

- Then for `count()` we would have:

```{r, eval=FALSE}
flights %>% count_(month, sort=TRUE)
```


--- .class #id

- Then for `count()` we would have:

```{r, echo=FALSE}
flights %>% count_(month, sort=TRUE)
```

--- .class #id

## Grouping with other functions

- We can also sum over other values rather than just counting the rows like the above examples. 
- For example let us say we were interested in knowing the total distance for planes in a given month. 
- We could do this with the `summarise()` function, `tally()` function or the `count()` function:

```{r, eval=FALSE}
flights %>% 
  group_by(month) %>% 
  summarise(dist = sum(distance))
```

--- .class #id

## Grouping with other functions

- We take flights then group by month and then create a new variable called distance, where we sum the distance. 
- For  `tally()` we could do:

```{r ,eval=FALSE}
flights %>% 
  group_by(month) %>% 
  tally(wt = distance)
```

*Note: in `tally()` the `wt` stands for weight and allows you to weight the sum based on the distance*. 

--- .class #id

## Grouping with other functions

- With the `count()` function we also use `wt`:


```{r}
flights %>% count(month, wt = distance)
```

--- .class #id

##Counting Groups


- We may want to know how large our groups are. To do this we can use the following functions:
  - `group_size()` is a function that returns counts of group. 
  - `n_groups()` returns the number of groups

--- .class #id

##Counting Groups

- So if wanted to count the number of flights by month, we could group by month and find the groups size using `group_size()`:

```{r, eval=FALSE}
flights %>% 
  group_by(month) %>% 
  group_size()
```


--- .class #id

##Counting Groups



```{r, echo=FALSE}
flights %>% 
  group_by(month) %>% 
  group_size()
```


--- .class #id

##Counting Groups


- If we just wished to know how many months were represented in our data we could use the `n_groups()` function:


```{r, eval=FALSE}
flights %>% 
  group_by(month) %>% 
  n_groups()
```



--- .class #id

##Counting Groups

```{r, echo=FALSE}
flights %>% 
  group_by(month) %>% 
  n_groups()
```







